Tackling discrepancies in freshwater invertebrate trait databases and taxonomic resolution: Harmonising across continents and trait aggregation

### 2021-01-27 Paper/Draft_harmonization_paper.tex
27.01.: Final edits upon sending to editor

% Intial: Between the trait databases, grouping features differed in terms of their classification into traits and their focus leading to discrepancies how traits are defined
3. We found that between trait databases, trait definitions are often different because varying numbers of traits are used to describe groups of related traits (e.g. respiration traits). Also the focus to describe groups of related traits varies between databases, e.g. for feeding mode some focus on the food source, while others on mouthpart morphology. Also, the coding to describe traits (binary, fuzzy) varied between databases. Aggregated and assigned traits based on expert knowledge displayed a good match, especially approaches using the median to aggregate from species to family-level. The comparison with the case study showed that harmonisation and trait aggregation only slightly changed results.

@@

Traits of individual freshwater invertebrate are difficult to determine because - unlike plant traits - they are often difficult to measure directly. For example, to describe feeding habits requires that we understand mouthpart morphology, feeding behavior, and the specific food that is consumed(\cite{moog_comprehensive_nodate}). Nevertheless, invertebrate traits have been increasingly used in freshwater ecology, e.g. by relating macroinvertebrate trait composition to environmental factors (\cite{bhowmik_large_2015, poff_developing_2010, szocs_effects_2014}).

@@

The definition of grouping features varied across databases mainly concerning their differentiation into traits but also in their focus. We provide a summary of discrepancies in trait definitions in the appendix (Table \ref{stab:trait_definitions}). Both, differences in differentiation and focus can lead to discrepancies in trait definitions. For example, for the grouping feature feeding mode discrepancies arise because traits are assigned in different ways. The Tachet database defines predators as carvers, engulfers and swallowers. By contrast, in the CONUS database predators are defined as engulfers and carnivorous piercers. In turn, in the Tachet database, piercers are defined as a separate trait encompassing herbivorous and carnivorous piercers. Furthermore, the focus in the freshwaterecology.info database for feeding mode is primarily on the food source of a species (except for filterers), while the other databases focus on the strategies of food acquisition. Therefore, the freshwaterecology.info database defines e.g. predator as "eating from prey", while the other databases use the mouthpart morphology as basis of their definition. The Tachet database captures the food source in an additional grouping feature. Varying levels of differentiation are also present in all other investigated grouping features between the trait databases (Table \ref{tab:trait_databases_coding_differentiation} and Table \ref{stab:trait_definitions}). Locomotion definitions differ also in focus between databases. Freshwaterecology.info and New Zealand databases describe locomotion as the way of movement of an organism, Tachet as substrate relation and locomotion, the Vieira as how organisms deal with flow, Australia as attachment, and the CONUS database includes not only the way of movement, but also the location of movement. Similarly, regarding reproduction traits, databases differ in their focus. Reproduction is captured in one grouping feature and defined as location of oviposit clutches and mode of reproduction in the freshwaterecology.info and Tachet databases. The Vieira database provides information on the oviposition location but not on reproductive behavior. The Australian database report traits for reproductive behavior but also on oviposition site. The New Zealand database distinguishes three grouping features related to reproduction: reproductive technique, oviposition site (e.g. water surface, terrestrial), and egg/egg mass (e.g. free, cemented).

@@

\subsection*{Differences in trait affinities obtained by trait aggregation methods compared to traits assigned at family-level}
\label{sec:diff_trait_agg_chessman}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=16.5cm, height=10cm]{Deviances_trait_agg_pyne.png}
  \caption{Cases (factor combination investigated families and traits) where differences occurred between aggregated traits and traits assigned at family-level. Violin plots - a mirrored density plot - show the density of the absolute trait affinity differences for the North America dataset for the grouping features feeding mode, locomotion, respiration, body size and voltinism. For more details see Figure \ref{fig:diff_aggr_traits_chessman}}
  \label{fig:diff_aggr_traits_pyne}
\end{figure}

@@

\subsection*{Re-analysis of Szöcs et al. using harmonised and aggregated grouping features}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=16.5cm, height=25cm]{Scores_on_constrained_axis_comb_ALT.pdf}
\end{figure}
\clearpage % insert a page break
\captionof{figure}{RDA of traits constrained by electric conductivity for the tested methods and the original study. Shown are boxplots of the site scores along the conductivity axis. The rug on the right side of each plot indicates species scores of the traits on the conductivity axis. Only traits with a mahalanobis distance greater than the $97.5 \%$ quantile of the Chi-square distribution (5.02) were labelled. For better comparability, species scores of the original analysis were multiplied by -1.} 
\label{fig:boxplots_scores_on_constrained_axis}

>>>>

% Initial: Between the trait databases, grouping features differed in terms of their classification into traits and their focus leading to discrepancies how traits are defined
3. We found that between trait databases, trait definitions are often different because varying numbers of traits are used to describe groups of related traits (e.g. respiration traits). Also the focus to describe groups of related traits varies between databases, e.g. for feeding mode some focus on the food source, while others on mouthpart morphology. Also, the coding to describe traits (binary, fuzzy) varied between databases. Aggregated and assigned traits based on expert knowledge displayed a good match, especially approaches using the median to aggregate from species to family-level. The comparison with the case study showed that harmonisation and trait aggregation only slightly changed results.

@@

Traits of individual freshwater invertebrate are difficult to determine because - unlike plant traits - they are often difficult to measure directly. For example, to describe feeding habits requires that we understand mouthpart morphology, feeding behaviour, and the specific food that is consumed (\cite{moog_comprehensive_nodate}). Nevertheless, invertebrate traits have been increasingly used in freshwater ecology, e.g. by relating macroinvertebrate trait composition to environmental factors (\cite{bhowmik_large_2015, poff_developing_2010, szocs_effects_2014}).

@@

The definition of grouping features varied across databases mainly concerning their differentiation into traits but also in their focus. We provide a summary of discrepancies in trait definitions in the appendix (Table \ref{stab:trait_definitions}). Both, differences in differentiation and focus can lead to discrepancies in trait definitions. For example, for the grouping feature feeding mode discrepancies arise because traits are assigned in different ways. The Tachet database defines predators as carvers, engulfers and swallowers. By contrast, in the CONUS database predators are defined as engulfers and carnivorous piercers. In turn, in the Tachet database, piercers are defined as a separate trait encompassing herbivorous and carnivorous piercers. Furthermore, the focus in the freshwaterecology.info database for feeding mode is primarily on the food source of a species (except for filterers), while the other databases focus on the strategies of food acquisition. Therefore, the freshwaterecology.info database defines e.g. predator as "eating from prey", while the other databases use the mouthpart morphology as basis of their definition. The Tachet database captures the food source in an additional grouping feature. Varying levels of differentiation are also present in all other investigated grouping features between the trait databases (Table \ref{tab:trait_databases_coding_differentiation} and Table \ref{stab:trait_definitions}). Locomotion definitions differ also in focus between databases. Freshwaterecology.info and New Zealand databases describe locomotion as the way of movement of an organism, Tachet as substrate relation and locomotion, the Vieira as how organisms deal with flow, Australia as attachment, and the CONUS database includes not only the way of movement, but also the location of movement. Similarly, regarding reproduction traits, databases differ in their focus. Reproduction is captured in one grouping feature and defined as location of oviposit clutches and mode of reproduction in the freshwaterecology.info and Tachet databases. The Vieira database provides information on the oviposition location but not on reproductive behaviour. The Australian database report traits for reproductive behaviour but also on oviposition site. The New Zealand database distinguishes three grouping features related to reproduction: reproductive technique, oviposition site (e.g. water surface, terrestrial), and egg/egg mass (e.g. free, cemented).

@@

\subsection*{Comparison of family-level aggregated traits with family-level assigned traits}
\label{sec:diff_trait_agg_chessman}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=16.5cm, height=10cm]{Deviances_trait_agg_pyne.png}
  \caption{Cases (factor combination investigated families and traits) where differences occurred between aggregated traits and traits assigned at family-level. Violin plots - a mirrored density plot - show the density of the absolute trait affinity differences for the North America dataset for the grouping features feeding mode, locomotion, respiration, body size and voltinism. For more details see Figure \ref{fig:diff_aggr_traits_chessman}.}
  \label{fig:diff_aggr_traits_pyne}
\end{figure}

@@

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=16.5cm, height=25cm]{Scores_on_constrained_axis_comb_ALT.pdf}
\end{figure}
\clearpage % insert a page break
\captionof{figure}{RDA of traits constrained by electric conductivity for the tested methods and the original study. Shown are boxplots of the site scores along the conductivity axis. The rug on the right side of each plot indicates species scores of the traits on the conductivity axis. Only traits with a Mahalanobis distance greater than the $97.5 \%$ quantile of the Chi-square distribution (5.02) were labelled. For better comparability, species scores of the original analysis were multiplied by -1.} 
\label{fig:boxplots_scores_on_constrained_axis}
